PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics

### 2018-05-31 draft.tex
Edited the energy conservation image - now 96^3 rather than 64^3 at lowest resolution.

The outcomes of ULDM soliton collisions depend critically on  whether the total energy of the isolated binary system is positive or negative. With a positive total energy the solitons pass through each other, emerging largely undisturbed from their initial configurations and the wavefunctions describing the solitons are superposed during the collision,, yielding distinctive interference patterns.

@@

As demonstrated in \cite{Paredes2016}, the wavelike properties of ULDM can give rise to effective forces which can dramatically affect the dynamics of core collisions. These effective forces arise as a result of interference phenomena, and do not require the ULDM model to incorporate explicit local interactions. Figure \ref{fig:repulsion} shows the results of two simulations using \PyUltraLight in which two solitons of equal mass undergo a head-on collision. Contours of the density profile along the plane of symmetry are displayed. In the first simulation, there is no phase offset between the initial solitons, while in the second simulation the phases differ by $\pi$. We can see that the result of this phase shift is to create an effective repulsive force between the two solitons as they approach one another. Further discussion of this phenomenon and its possible observational consequences can be found in \cite{Paredes2016}.
\begin{figure}
  \includegraphics[width=1.\textwidth, trim={0 0 0 0},clip]{phase_comparison}
  \caption{Head-on collisions between two solitons of mass 5, relative velocity 6, and initial separation 4 in code units. Time progresses from left to right across each row.}
  \label{fig:repulsion}
\end{figure}

@@

@@

When the SP system is recast as a system of hydrodynamical equations, the fluid velocity is given by the gradient of the smoothly-varying phase at any point. This implies that the velocity flow of the SP system is irrotational, $\nabla\times\vec{v}=0$. When a soliton is initialised in a circular orbit around a Newtonian potential, the initially spherical profile is distorted, elongated along the radial direction of the central potential. Meanwhile, the velocity field corresponding to the overall orbital motion of the soliton is superposed with the internal velocity field, combining so as to produce a net flow with vanishing curl. The family of Riemann-S ellipsoids describe uniformly rotating bodies which satisfy the condition of irrotationality \cite{Chandrasekhar1965}, and it is therefore the characteristic internal velocity field of such an ellipsoid which we expect to arise during the simulation. It is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked irrotational Riemann-S ellipsoid when orbiting a host mass. We are able to reproduce this result using \PyUltraLight, with Figure \ref{fig:riemann} showing the internal velocity field of a solitonic satellite after one complete revolution around a host mass, obtained using \PyUltraLight on a $256^3$ simulation grid. In this case the ratio of host to satellite mass was $\approx 55$. It can be seen that the soliton has become elongated along the radial line connecting to the host, indicating that it is tidally locked. Meanwhile the internal velocity field can be seen to be that of an irrotational Riemann-S ellipsoid \cite{Daller2012}.
\begin{figure}
  \includegraphics[width=1.\textwidth,trim=1cm 1cm 0 1cm,clip]{riemann}
  \caption{Solitonic core after one revolution of a central potential, demonstrating the internal velocity field of an irrotational Riemann-S ellipsoid (where the collective motion of the soliton has been subtracted from the velocity field). Deformation of the spherical density profile into that of an ellipsoid can be seen, with extension along the radial direction (red arrow).}
  \label{fig:riemann}
\end{figure}

@@

@@

Figures \ref{fig:combined_1} and \ref{fig:binary} demonstrate energy conservation for two example simulation scenarios. In the first case, we examine the evolution of the energy for a single soliton undergoing significant tidal disruption within a Newtonian central potential. We see that as the soliton is disrupted, the kinetic energy increases, while the gravitational energy due to the central potential decreases, as expected. Meanwhile, the gravitational potential energy due to self-interaction gradually increases toward zero as the disruption continues. The sum of the individual energy components is conserved to $10^{-2}\ \%$ in this case, where the simulation was run at $256^3$. Higher grid resolutions result in even better energy conservation. Figure \ref{fig:binary} demonstrates the evolution of the energy of a binary system of solitons in elliptical orbits around their common centre of mass. In this case the simulation ran for approximately three orbital periods at $256^3$. At points of closest approach, it can be clearly seen that the kinetic energy increases as the solitons speed up, while the potential energy due to self-interaction decreases commensurately. In this scenario no central potential has been included, so the gravitational energy due to the central potential stays uniformly zero throughout the simulation and is not plotted in the Figure. It can also be seen that as the solitons reach the first point of closest approach, they become slightly deformed, exciting oscillatory modes which are manifest in the Figure as small scale oscillations superposed on the global behaviour. In this case the total energy is conserved to $10^{-3}\ \%$.

% \begin{figure}[tbp]
% \centering 
% \includegraphics[width=1.\textwidth,trim=0 0 0 0,clip]{density_diagram1}
% \caption{Density profile of solitonic core at equally spaced times as it undergoes tidal disruption in a potential centred at the red cross.}\label{fig:density_diagram1}
% \end{figure}

% \begin{figure}[tbp]
% \centering 
% \includegraphics[width=1.\textwidth,trim=0 0 0 0,clip]{energy_diagram1}
% \caption{Evolution of the energy of the system illustrated in Figure \ref{fig:density_diagram1}. Times are indicated in correspondence with the snapshots of the density profile. }\label{fig:energy_diagram_1}
% \end{figure}

% \begin{figure}[tbp]
% \centering 
% \includegraphics[width=1.1\textwidth,trim=2.5cm 0 0 1cm,clip]{combined_energy_and_density1}
% \caption{Left: Evolution of the density profile of a solitonic core at equally spaced times as it undergoes tidal disruption in a potential centred at the red cross. Right: Evolution of the energy of the system, where times are indicated in correspondence with the snapshots of the density profile. All quantities are in dimensionless code units.}\label{fig:combined_1}
% \end{figure}

% \begin{figure}[tbp]
% \centering 
% \includegraphics[width=1.\textwidth,trim=0 0 2cm 2cm,clip]{energy_diagram2}
% \caption{Evolution of the energy for a binary soliton system with each soliton in an elliptical orbit around the common centre of mass.}\label{fig:binary}
% \end{figure}

>>>>

The outcomes of ULDM soliton collisions depend critically on  whether the total energy of the isolated binary system is positive or negative. With a positive total energy the solitons pass through each other, emerging largely undisturbed from their initial configurations and the wavefunctions describing the solitons are superposed during the collision, yielding distinctive interference patterns.

@@

\begin{figure}
  \includegraphics[width=1.\textwidth, trim={0 0 0 0},clip]{phase_comparison}
  \caption{Head-on collisions between two solitons of mass 5, relative velocity 6, and initial separation 4 in code units. Time progresses from left to right across each row.}
  \label{fig:repulsion}
\end{figure}

As demonstrated in \cite{Paredes2016}, the wavelike properties of ULDM can give rise to effective forces which can dramatically affect the dynamics of core collisions. These effective forces arise as a result of interference phenomena, and do not require the ULDM model to incorporate explicit local interactions. Figure \ref{fig:repulsion} shows the results of two simulations using \PyUltraLight in which two solitons of equal mass undergo a head-on collision. Contours of the density profile along the plane of symmetry are displayed. In the first simulation, there is no phase offset between the initial solitons, while in the second simulation the phases differ by $\pi$. We can see that the result of this phase shift is to create an effective repulsive force between the two solitons as they approach one another. Further discussion of this phenomenon and its possible observational consequences can be found in \cite{Paredes2016}.

@@

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=1cm 1cm 0 1cm,clip]{riemann}
  \caption{Solitonic core after one revolution of a central potential, demonstrating the internal velocity field of an irrotational Riemann-S ellipsoid (where the collective motion of the soliton has been subtracted from the velocity field). Deformation of the spherical density profile into that of an ellipsoid can be seen, with extension along the radial direction (red arrow).}
  \label{fig:riemann}
\end{figure}

@@

When the SP system is recast as a system of hydrodynamical equations, the fluid velocity is given by the gradient of the smoothly-varying phase at any point. This implies that the velocity flow of the SP system is irrotational, $\nabla\times\vec{v}=0$. When a soliton is initialised in a circular orbit around a Newtonian potential, the initially spherical profile is distorted, elongated along the radial direction of the central potential. Meanwhile, the velocity field corresponding to the overall orbital motion of the soliton is superposed with the internal velocity field, combining so as to produce a net flow with vanishing curl. The family of Riemann-S ellipsoids describe uniformly rotating bodies which satisfy the condition of irrotationality \cite{Chandrasekhar1965}, and it is therefore the characteristic internal velocity field of such an ellipsoid which we expect to arise during the simulation. It is found in \cite{Du2018} that an initially spherical solitonic core without self-rotation will gradually spin up to form a tidally-locked irrotational Riemann-S ellipsoid when orbiting a host mass. We are able to reproduce this result using \PyUltraLight, with Figure \ref{fig:riemann} showing the internal velocity field of a solitonic satellite after one complete revolution around a host mass, obtained using \PyUltraLight on a $256^3$ simulation grid. In this case the ratio of host to satellite mass was $\approx 55$. It can be seen that the soliton has become elongated along the radial line connecting to the host, indicating that it is tidally locked. Meanwhile the internal velocity field can be seen to be that of an irrotational Riemann-S ellipsoid \cite{Daller2012}.

%\vspace{1em}

@@

\begin{figure}
  \includegraphics[width=0.9\textwidth,trim=0 0.5cm 0 0,clip]{Total_Energy_Change.eps}
  \caption{Demonstration of improving energy conservation with simulation resolution for a binary system of solitons orbiting their common centre of mass.}
  \label{fig:energy_change}
\end{figure}

@@

Figures \ref{fig:combined_1} and \ref{fig:binary} demonstrate energy conservation for two example simulation scenarios. In the first case, we examine the evolution of the energy for a single soliton undergoing significant tidal disruption within a Newtonian central potential. For this simulation a soliton of mass 12 in code units was initialised at a radial distance of 3 code units from the centre of a Newtonian central potential generated by a central mass of 1000 code units. We see that as the soliton is disrupted, the kinetic energy increases, while the gravitational energy due to the central potential decreases, as expected. Meanwhile, the gravitational potential energy due to self-interaction gradually increases toward zero as the disruption continues. The sum of the individual energy components is conserved to $10^{-2}\ \%$ in this case, where the simulation was run at $256^3$. Higher grid resolutions result in even better energy conservation.

Figure \ref{fig:binary} demonstrates the evolution of the energy of a binary system of solitons in elliptical orbits around their common centre of mass. In this case the simulation ran for approximately three orbital periods at $256^3$. In dimensionless code units, the soliton mass was 22, the initial separation was 2, and the initial relative velocity was 3.6. At points of closest approach, it can be clearly seen that the kinetic energy increases as the solitons speed up, while the potential energy due to self-interaction decreases commensurately such that the total energy is conserved. In this scenario no central potential has been included, so the gravitational energy due to the central potential stays uniformly zero throughout the simulation and is not plotted in the Figure. It can also be seen that as the solitons reach the first point of closest approach, they become slightly deformed, exciting oscillatory modes which are manifest in the Figure as small scale oscillations superposed on the global behaviour. In this case the total energy is conserved to $10^{-3}\ \%$.

Figure \ref{fig:energy_change} demonstrates the relationship between the total integrated energy and the grid resolution for the same binary system of solitons used to generate Figure \ref{fig:binary}. While energy is conserved at sub-percent level even for low resolutions ($96^3$), this is dramatically improved at higher grid resolutions.
